How to select predictive models for decision-making or causal inference?

### 2024-11-10 response_to_reviewer.md

*I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.*

*Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.*

> We thank the reviewer for the thoughful comments and the call for
> nuance. Our first response is that our thoughts did not come out right
> and that it is not clear, indeed, that R-risk is always the best.
> Rather, that it is a good default solution. We have reformulate parts of
> the paper to convey this nuance better and added the nuance in the
> conclusion.

> In addition, the reviewer is correct that factors beyond overlap
> probably affect which risk is best. We have followed the suggestion to
> explore more diverse simulation scenarii, as detailing the sensitivity
> of our result on the R-risk makes a more nuanced and stronger analysis.
> Thus, we ran a dedicated experiment focusing on the ratio of causal
> effect to background response. We measure the ratio by the absolute
> mean difference between the causal effect and the baseline response,
> measured empirically on each dataset instance $\Delta_{\mu} =
> \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) -
> \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_{0}(x_i) + \mu_{1}(x_i)} \big|$. We vary a dedicated parameter
> ($\omega$ in Algorithm 2) in the simulation to force this ratio to
> cover a wide range of values. In 90 dataset instance, it ranges several
> orders of magnitude from 0.04 to 206. Figure 7 gives the main results
> of this experiment. We see that R-risk is still the best risk on
> average, but when the effect ratio is high (tertile between 9 and
> 9675), the reweighted mu-risk and the tau-risk_IPW are better than
> the R-risk. However, for low value of the ratio,  the R-risk largely
> outperforms all other metrics. We added these new results to the main
> paper. We also studied another characteristic of the simulation: the variance of the causal effect measured as the empirical version of $\frac{var(\mu_1 - \mu_0)}{var(\mu_1+\mu_0)}$. For this last experiment, we found similar results to the causal effect experiment. We include it in the supp mat for completeness about our reflexion.

### 2024-11-10 how_to_select_latex.tex
iter submission of revised manuscript

Beyond overlap, we study for caussim simulations, the effect on model selection
of different causal effect ratio to baseline. We vary the empirical mean
absolute difference between the causal effect and the baseline, $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$, covering a ratio range from 0.04 to 206 ($\text{median}=9.1$, full
distribution in Appendix \ref{apd:effect_ratio}). Figure
\ref{fig:effect_ratio_influence} shows that for high values of the ratio,
$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
better risk.

@@

Going beyond overlap, we study the effect of another important parameter of the
data generation process: the ratio of causal effect to the baseline response.
Here, we detail the experimental setup of the results presented in Figure \ref{fig:effect_ratio_influence}.

@@

>>>>

Beyond overlap, we study for caussim simulations, the effect on model selection
of different causal effect ratio to baseline. We vary the empirical mean
absolute difference between the causal effect and the baseline, $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$, covering a ratio range from 0.04 to 206 ($\text{median}=9.1$). Appendix \ref{apd:effect_ratio} details this setup as well as an alternive measure of effect ratio. Figure
\ref{fig:effect_ratio_influence} shows that for high values of the ratio,
$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
better risk.

@@

Going beyond overlap, we study the effect of another important parameter of the
data generation process: the ratio of causal effect to the baseline response.

Here, we detail the experimental setup of the results presented in Figure \ref{fig:effect_ratio_influence}.

@@

We also studied another measure, focusing on the ratio of the variance of the causal effect to the variance of the mean shared response ($\mu_0 + \mu_1$) of both population: $\frac{\hat{var}(\mu_{1}(x_i) - \mu_{0}(x_i))}{\hat{var}(\mu_{1}(x_i) + \mu_{0}(x_i))}$, covering a range from 0 to 11.08 with the full
distribution described in Table \ref{table:effect_variance}. Figure
\ref{fig:effect_variance_influence} shows that for high values of the variance,
$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
better risk.

\begin{figure}[!tb]
    \centering
    \begin{minipage}{\linewidth}
        \includegraphics[width=0.9\linewidth]{_6_effect_variance_by_bin_comparaison_kendall_by_Dataset.pdf}
    \end{minipage}

    \caption{\textbf{R-risk is robust to a wide range of effect variance}:
    Kendall's $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal
    effect variance are the tertiles of the ratio of causal effect variance to
    the mean outcome variance: Low $[0;0.04[$, Medium $[0.04; 1.04[$,
    High $[1.04;11.08[$.}\label{fig:effect_variance_influence}
\end{figure}

\begin{table}[!ht]
    \centering
    \resizebox{\textwidth}{!}{ % Adjusts the table to fit the text width
        \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
            \hline
                                      & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
            \textbf{effect\_variance} & 179            & 1.013         & 1.781        & 1.680e-5     & 1.734e-5     & 4.875e-4      & 6.449e-3      & 0.789         & 1.160         & 1.807         & 9.641         & 11.079       \\
            \hline
        \end{tabular}
    }
    \caption{Statistical summary of the causal effect variance.}
    \label{table:effect_variance}
\end{table}
